---
title: "Description and usage of MsBackendMassbank"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Description and usage of Description and usage of MsBackendMassbank}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{Spectra}
    %\VignetteDepends{Spectra,BiocStyle}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("MsBackendMassbank")`<br />
**Authors**: `r packageDescription("MsBackendMassbank")[["Author"]] `<br />
**Last modified:** `r file.info("MsBackendMassbank.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r, echo = FALSE, message = FALSE}
library(Spectra)
library(BiocStyle)
```

# Introduction

The `Spectra` package provides a central infrastructure for the
handling of Mass Spectrometry (MS) data. The package supports
interchangeable use of different *backends* to import MS data from a
variety of sources (such as mzML files). The `MsBackendMassbank` package
allows the import of handle MS/MS spectrum 
data from [Massbank](https://github.com/MassBank/MassBank-data) files. 
This vignette illustrates the usage of the `MsBackendMassbank`
package.

# Importing MS/MS data from MassBank files

MassBAnk files store normally one library spectrum per file, typically
centroided and of MS level 2. In our short example below, we load data from a 
file containing multiple library spectra per file or from files with each a 
single spectrum provided with this package. Below we first load all required packages
and define the paths to the Massbank files.

```{r load-libs}
library(Spectra)
library(MsBackendMassbank)
fls <- dir(system.file("extdata", package = "MsBackendMassbank"),
           full.names = TRUE, pattern = "txt$")
fls
```

MS data can be accessed and analyzed through `Spectra` objects. Below
we create a `Spectra` with the data from these mgf files. To this end
we provide the file names and specify to use a `MsBackendMassbank()`
backend as *source* to enable data import. First we import from a single file
with multiple library spectra.

```{r import}
sps <- Spectra(fls[1], source = MsBackendMassbank())
```

With that we have now full access to all imported spectra variables
that we list below.

```{r spectravars}
spectraVariables(sps)
```

The same is possible with multiple files, containing each one individual library
spectrum.

```{r import2}
sps <- Spectra(fls[-1], source = MsBackendMassbank())
spectraVariables(sps)
```

By default the complete metadata is read together with the spectra. This can
increase loading time. The different metadata blocks can blocked from reading.
This requires to define an additional data frame indicating what shall be read.

```{r metadata}
# create data frame to indicate with metadata blocks shall be read.
metaDataBlocks <- data.frame(metadata = c("ac", "ch", "sp", "ms", "record", "pk", "comment"),
                             read = rep(FALSE, 7),
                             stringsAsFactors = FALSE)

sps <- Spectra(fls[-1], source = MsBackendMassbank(), metaBlock = metaDataBlocks)

# variables only contain core variables
spectraVariables(sps)
```


Besides default spectra variables, such as `msLevel`, `rtime`,
`precursorMz`, we also have additional spectra variables such as the
`title` of each spectrum in the mgf file.

```{r instrument}
sps$rtime
sps$title
```

In addition we can also access the m/z and intensity values of each
spectrum.

```{r mz}
mz(sps)
intensity(sps)
```

When importing a large number of mgf files, setting `nonStop = TRUE`
prevents the call to stop whenever problematic mgf files are
encountered. 

```{r all-import, eval = FALSE}
sps <- Spectra(fls, source = MsBackendMgf(), nonStop = TRUE)
```

# Session information

```{r}
sessionInfo()
```
